Constructing a universe for the setoid model

### 2020-10-13 paper.tex
Fix citations

Intuitionistic type theory is a formal system designed by Per Martin-L\"of to be
a full-fledged foundation in which to develop constructive mathematics
\cite{mltt75, libretto}.
%
A central aspect of type theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand ``propositional'' equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics, one should aim for a notion of
propositional equality that is as \emph{extensional} as possible.

@@

This question was revisited and answered in Altenkirch et al. \cite{mpc19}. In
this paper, the authors define Setoid Type Theory (SeTT), an extension of
intensional Martin-L\"of type theory with constructs for setoid reasoning, where
funext and propext hold by definition. SeTT is based on the \emph{strict} setoid
model of Altenkirch\footnote{a \emph{strict} model is one where every equation
  holds definitionally}, which makes it possible to show consistency via a
syntactic translation. This is in contrast with other type theories based on the
setoid model, like Observational Type Theory \cite{alti:ott-conf, alti:ott-conf}
and XTT \cite{xtt}, which instead rely on ETT for their justification. A major
property of SeTT is thus to illustrate how to bootstrap extensionality, by
translation into a small intensional core.

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf, alti:ott-conf} is a syntax for the setoid
model differing from SeTT for the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.
%
Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
universe using a translation into constructive set theory. In
contrast, we target a basic intensional type theory and hence also
give a computational semantics.

@@

A known and established method for defining universes in type theory relies on
induction-recursion (IR), a definition schema developed by Dybjer
\cite{ir-dybjer, dybjer-setzer}. Inductive-recursive definitions can be found
throughout the literature, from the already mentioned type theoretic universes,
including the original formulation \`{a} la Tarski by Martin-L\"of
\cite{libretto}, to metamathematical tools like computability predicates.

>>>>

Intuitionistic type theory is a formal system designed by Per Martin-L\"of to be
a full-fledged foundation in which to develop constructive mathematics
\cite{mltt75,libretto}.
%
A central aspect of type theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand ``propositional'' equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics, one should aim for a notion of
propositional equality that is as \emph{extensional} as possible.

@@

This question was revisited and answered in Altenkirch et al. \cite{mpc19}. In
this paper, the authors define Setoid Type Theory (SeTT), an extension of
intensional Martin-L\"of type theory with constructs for setoid reasoning, where
funext and propext hold by definition. SeTT is based on the \emph{strict} setoid
model of Altenkirch\footnote{a \emph{strict} model is one where every equation
  holds definitionally}, which makes it possible to show consistency via a
syntactic translation. This is in contrast with other type theories based on the
setoid model, like Observational Type Theory \cite{alti:ott-conf,alti:ott-conf}
and XTT \cite{xtt}, which instead rely on ETT for their justification. A major
property of SeTT is thus to illustrate how to bootstrap extensionality, by
translation into a small intensional core.

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf,alti:ott-conf} is a syntax for the setoid
model differing from SeTT for the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.
%
Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
universe using a translation into constructive set theory. In
contrast, we target a basic intensional type theory and hence also
give a computational semantics.

@@

A known and established method for defining universes in type theory relies on
induction-recursion (IR), a definition schema developed by Dybjer
\cite{ir-dybjer,dybjer-setzer}. Inductive-recursive definitions can be found
throughout the literature, from the already mentioned type theoretic universes,
including the original formulation \`{a} la Tarski by Martin-L\"of
\cite{libretto}, to metamathematical tools like computability predicates.
